A Survey on Ordered Binary Decision Diagrams

### 2015-09-15 paper.tex
Revise Symbolic model checking

One specific application of reduced ordered binary decision diagrams is 
\textit{symbolic model checking}, introduced by Burch et al.~\cite{BURCH90} in
1990. \textit{Symbolic}, because the state space is not represented explicitly
but symbolically by utilizing OBDDs to represent $\mu$-Calculus formulas.
In~\cite{BURCH90}, Burch, Clarke, McMillan, Dill and Hwang derived an efficient
decision procedures for CTL model checking as well as satisfiability of LTL
formulas, strong and weak observational equivalence of finite transition systems
and language containment for finite $\omega$-automata by introducing a new
$\mu$-Calculus model checking algorithm. However, this survey just outlines the
$\mu$-Calculus and the efficient decision procedures for CTL model checking.

The main idea behind this approach is to represent the state space symbolically,
instead of explicitly with e.g. a state table. Such explicit state tables may grow
exponentially with the number of concurrent components, because the state space
does so. Commonly, this is referred to as the
\textit{state explosion problem}~\cite{BURCH90, MCMILLAN92}. To overcome this
problem, Burch et al.~\cite{BURCH90} as well as McMillan~\cite{MCMILLAN92}
proposed to utilize reduced ordered binary decision diagrams to represent the 
state space symbolically. Although OBDDs cannot avoid a state explosion in general,
many practically relevant systemes with large state spaces can be
verified~\cite{BURCH90, MCMILLAN92}. Experiments~\cite{BURCH90} showed that
symbolic model checking enables to verify models with more than $10^{20}$ states,
whereas explicit enumeration approaches can only verify models up to $10^6$
states.

Burch et al.~\cite{BURCH90} and McMillan~\cite{MCMILLAN92} extend Bryant's OBDD
representation~\cite{BRYANT86} by quantification over Boolean variables (QBF) and
substitution of variable names. The quantification over Boolean variables is
realized by using Bryant's \textit{Restrict} operation (see
Section~\ref{sec:operations}). Let $f$ be a QBF formula and $x$ be a Boolean
variable. The QBF formula $\exists x\left(f\right)$ then is equivalent to
$f|_{a=0} \lor f|_{a=1}$~\cite{BURCH90}. Substitution of variables, e.g.
substitution of $x$ by $y$ (denoted $f\langle x\leftarrow y\rangle$), is computed
using the following equivalence:
\begin{center}
$
f\langle x\leftarrow y\rangle \equiv
\exists x\lbrack\left(x \Leftrightarrow y\right) \land f\rbrack
$
\end{center}

Before discussing the symbolic model checking algorithms $\text{BDD}_f$ and
$\text{BDD}_R$, the used $\mu$-Calculus has to be described. For a given finite
signature $\mathcal{S}$, each symbol is either an individual or a predicate
variable having positive arity. Burch et al.~\cite{BURCH90} considered two
syntactic categories, i.e. \textit{formulas} and \textit{relational terms}, having
the following forms.

\noindent
\textbf{Formulas:}~\cite{BURCH90}
\begin{enumerate}
    \item{\textbf{True}, \textbf{False}}.
    \item{
        $\lbrack z_1 = z_2 \rbrack$; $z_{1, 2}\in\mathcal{S}$ are individual
        variables.
    }
    \item{
        $\neg f$, $f \lor g$, $\exists z\left( f \right)$; $z\in\mathcal{S}$ is
        an individual variable, $f$ and $g$ are formulas.
    }
    \item{
        $P\left(z_1, \ldots, z_n\right)$; $P$ is an $n$-ary relational term and
        \newline $z_{1, \ldots, n}\in\mathcal{S}$ are individual variables not
        free in $P$.
    }
\end{enumerate}
\textbf{$n$-ary relational terms:}~\cite{BURCH90}
\begin{enumerate}
    \item{$Z$; $Z\in\mathcal{S}$ is an $n$-ary predicate variable.}
    \item{
        $\lambda z_1, \ldots, z_n\left( f \right)$; $f$ is a formula and
        $z_1, \ldots, z_n\in\mathcal{S}$ are distinct individual variables.
    }
    \item{
        $\mu Z \left( P \right)$; $Z\in\mathcal{S}$ is an $n$-ary predicate variable
        and $P$ is an $n$-ary relational term \textit{formally monotone in} $Z$,
        i.e. all free occurrences of $Z$ in $P$ have to be evenly negated. This
        $n$-ary relatioinal term is the \textit{least fixed point} (\textit{lfp})
        of an $n$-ary relational term $P$. Analogous, $\nu Z \left( P \right)$ is
        the \textit{greatest fixed point} (\textit{gfp}) and is equivalent to
        $\neg\mu Z \left( \neg P \langle Z \leftarrow \left( \neg Z \right) \rangle \right)$.
    }
\end{enumerate}

@@

Finally, after defining the $\mu$-Calculus formally, the model checking algorithms
can be described. They operate on the limited Boolean domain, i.e.
$D = \lbrace 0, 1 \rbrace$ but Burch et al.~\cite{BURCH90} also described how any
\textit{finite} domain can be encoded into this limited Boolean domain. This last
fact generalizes the introduced algorithms to any \textit{finite} domain.

They described two function representing the algorithm:
$\text{BDD}_f \left( f, I_P \right)$ and $\text{BDD}_R \left( R, I_P \right)$,
where $f$ and $R$ denote a formula and a relational term, respectively. For 
$\text{BDD}_f$, $I_P$ represents a relational variable interpretation assigning
values to free relational variables in $f$~\cite{BURCH90}.

A given individual variable interpretation $I_D$ satisfies \newline
$\text{BDD}_f \left( f, I_P \right)$ iff $f$ is satisfied by $\mathcal{M}$ having
interpretations $I_P$ and $I_D$. OBDDs in combination with place-holder variables
$d_1, \ldots, d_n$ are used to represent the values of the relational variables
$\in I_P$. $d_i$ represents the $i^{th}$ argument of a relation and an $n$-ary
relation holds for arguments $y_1, \ldots, y_n$ iff the interpretation
$\langle d_1 \leftarrow y_1, \ldots, d_n \leftarrow y_n \rangle$ satisfies the
OBDD.

In essence, $\text{BDD}_f \left( f, I_P \right)$ distinguishes between five cases
derived from the OBDD semantics and the $\mu$-Calculus, utilizing the graph-based
algorithms of Bryant~\cite{BRYANT86} (also see Section~\ref{sec:operations}) as
well as the $\text{BDD}_R$ function (for applying a relational term $R$) with a
subsequent substitution of the place-holder variables with the argument variables, 
i.e. \newline
$\langle d_1 \leftarrow x_1, \ldots, d_n \leftarrow x_n \rangle$~\cite{BURCH90}.

@@

$\text{\textbf{EX}} f$ is satisfied for a state $s$ iff a state satisfying $f$
exists and $R \left( s, t \right)$ holds. In terms of $\mu$-Calculus relational
terms, $\text{\textbf{EX}} f$ can be formulated as
$\lambda u \lbrack \exists v \left( f(v) \land R \left( u, v \right) \right) \rbrack$~\cite{BURCH90}.

@@

Utilizing these equivalent representations, one efficient CTL model checking.
Moreover, Burch, Clarke, McMillan and Dill~\cite{BURCH90-2} described how to use
these methods to do \textit{fair} CTL model checking. \tbr

>>>>

One specific application of reduced ordered binary decision diagrams is 
\textit{symbolic model checking}, introduced by Burch et al.~\cite{BURCH90} in
1990. Symbolic, because the state space is not represented explicitly but
symbolically by utilizing OBDDs to represent $\mu$-Calculus formulas.
In~\cite{BURCH90}, Burch, Clarke, McMillan, Dill and Hwang derived efficient
decision procedures for CTL model checking as well as satisfiability of LTL
formulas, strong and weak observational equivalence of finite transition systems
and language containment for finite $\omega$-automata by introducing a new
$\mu$-Calculus model checking algorithm. However, this survey just outlines the
$\mu$-Calculus and the efficient decision procedures for CTL model checking.

The main idea behind this approach is to represent the state space symbolically,
instead of explicitly with e.g. a state table. Such explicit state tables may grow
exponentially with the number of concurrent components, because the state space
does so. Commonly, this is referred to as the
\textit{state explosion problem}~\cite{BURCH90, MCMILLAN92}. To overcome this
problem, Burch et al.~\cite{BURCH90} as well as McMillan~\cite{MCMILLAN92}
proposed to utilize reduced ordered binary decision diagrams to represent the 
state space symbolically. Although OBDDs cannot avoid a state explosion in general,
many practically relevant systems with large state spaces can be
verified~\cite{BURCH90, MCMILLAN92}. Experiments~\cite{BURCH90} showed that
symbolic model checking enables to verify models with more than $10^{20}$ states,
whereas explicit enumeration approaches can only verify models up to $10^6$
states.

Burch et al.~\cite{BURCH90} and McMillan~\cite{MCMILLAN92} extended Bryant's OBDD
representation~\cite{BRYANT86} by quantification over Boolean variables (QBF) and
substitution of variable names. The quantification over Boolean variables is
realized by using Bryant's \textit{Restrict} operation (see
Section~\ref{sec:operations}). Let $f$ be a QBF formula and $x$ be a Boolean
variable. The QBF formula $\exists x\left(f\right)$ then is equivalent to
$f|_{a=0} \lor f|_{a=1}$~\cite{BURCH90}. Substitution of variables, e.g.
substitution of $x$ by $y$ (denoted $f\langle x\leftarrow y\rangle$), is computed
using the following definition:
\begin{center}
$
f\langle x\leftarrow y\rangle :=
\exists x\lbrack\left(x \Leftrightarrow y\right) \land f\rbrack
$
\end{center}

Before discussing the symbolic model checking algorithm, the used $\mu$-Calculus
has to be defined. For a given finite signature $\mathcal{S}$, each symbol is
either an individual or a predicate variable having positive arity. Burch et
al.~\cite{BURCH90} considered two syntactic categories, i.e. \textit{formulas}
and \textit{relational terms}, having the following forms.

\vskip5pt\noindent
\textbf{Formulas:}~\cite{BURCH90}
\begin{enumerate}
    \item{\textbf{True}, \textbf{False}}.
    \item{
        $\lbrack z_1 = z_2 \rbrack$; $z_{1, 2}\in\mathcal{S}$ are individual
        variables.
    }
    \item{
        $\neg f$, $f_1 \lor f_2$, $\exists z\left( f \right)$; $z\in\mathcal{S}$ is
        an individual variable, $f_1$ and $f_2$ are formulas.
    }
    \item{
        $P\left(z_1, \ldots, z_n\right)$; $P$ is an $n$-ary relational term and
        \newline $z_{1, \ldots, n}\in\mathcal{S}$ are individual variables not
        free in $P$.
    }
\end{enumerate}
\textbf{$n$-ary relational terms:}~\cite{BURCH90}
\begin{enumerate}
    \item{$Z$; $Z\in\mathcal{S}$ is an $n$-ary predicate variable.}
    \item{
        $\lambda z_1, \ldots, z_n\left( f \right)$; $f$ is a formula and
        $z_1, \ldots, z_n\in\mathcal{S}$ are distinct individual variables.
    }
    \item{
        $\mu Z \left( P \right)$; $Z\in\mathcal{S}$ is an $n$-ary predicate variable
        and $P$ is an $n$-ary relational term \textit{formally monotone in} $Z$,
        i.e. all free occurrences of $Z$ in $P$ have to be evenly negated. This
        $n$-ary relatioinal term is the \textit{least fixed point} (\textit{lfp})
        of an $n$-ary relational term $P$. Analogous, $\nu Z \left( P \right)$ is
        the \textit{greatest fixed point} (\textit{gfp}) and is equivalent to
        $\neg\mu Z \left( \neg P \langle Z \leftarrow \left( \neg Z \right) \rangle \right)$.
    }
\end{enumerate}

@@

Finally, after defining the $\mu$-Calculus formally, the model checking algorithm
can be described, which operates on the limited Boolean domain, i.e.
$D = \lbrace 0, 1 \rbrace$. However, Burch et al.~\cite{BURCH90} also described
how any \textit{finite} domain can be encoded into this limited Boolean domain.
This last fact generalizes the introduced algorithm to any \textit{finite} domain.

They described two functions representing the algorithm:
$\text{BDD}_f \left( f, I_P \right)$ and $\text{BDD}_R \left( R, I_P \right)$,
where $f$ and $R$ denote a formula and a relational term, respectively. For 
$\text{BDD}_f$, $I_P$ represents a relational variable interpretation assigning
values to free relational variables in $f$~\cite{BURCH90}.

A given individual variable interpretation $I_D$ satisfies \newline
$\text{BDD}_f \left( f, I_P \right)$ iff $f$ is satisfied by $\mathcal{M}$ having
interpretations $I_P$ and $I_D$. OBDDs in combination with place-holder variables
$d_1, \ldots, d_n$ are used to represent the values of the relational variables
in $I_P$. $d_i$ represents the $i^{th}$ argument of a relation and an $n$-ary
relation holds for arguments $y_1, \ldots, y_n$ iff the interpretation
$\langle d_1 \leftarrow y_1, \ldots, d_n \leftarrow y_n \rangle$ satisfies the
OBDD.

In essence, $\text{BDD}_f \left( f, I_P \right)$ distinguishes between five cases
derived from the OBDD semantics and the $\mu$-Calculus, utilizing the graph-based
algorithms of Bryant~\cite{BRYANT86} (see Section~\ref{sec:operations}) as
well as the $\text{BDD}_R$ function (for applying a relational term $R$) with a
subsequent substitution of the place-holder variables with the argument variables, 
i.e. \newline
$\langle d_1 \leftarrow x_1, \ldots, d_n \leftarrow x_n \rangle$~\cite{BURCH90}.

@@

$\text{\textbf{EX}} f$ is satisfied for a state $s$ iff a state $t$ satisfying
$f$ exists and $R \left( s, t \right)$ holds. In terms of $\mu$-Calculus
relational terms, $\text{\textbf{EX}} f$ can be formulated as
$\lambda u \lbrack \exists v \left( f(v) \land R \left( u, v \right) \right) \rbrack$~\cite{BURCH90}.

@@

Utilizing these equivalent representations, efficient CTL model checking can be
done. Moreover, Burch, Clarke, McMillan and Dill~\cite{BURCH90-2} described how
to use these methods to do \textit{fair} CTL model checking.
